Determining the Structure of Length-$k$ Steenrod Operations as $\A(r)$-Modules

### 2015-06-18 paper.tex
maketitle is useful I hear
Add pictures, some other updates

\title{Determining the Structure of Length-$k$ Steenrod Operations as $\A(r)$-Modules}
\author{Benjamin Kraft}

@@

\begin{abstract}
  The Steenrod Algebra $\A$ is the algebra of stable natural endomorphisms of the $\mathbb{Z}/2$-cohomology functor; it is generated by elements $Sq^{2^i}$.  Let $\A(k)$ be the subalgebra generated by the $Sq^{2^i}$ for $i\leq k$.  Consider the modules $L(k)$ spanned by sequences of Steenrod operations of length $k$.  Welcher proved that $L(k)$ is a free module over $\A(k-1)$.  We are interested in finding the structure of $L(k)$ as an $\A(r)$-module for any $r$.  We conjecture that $L(k)$ is built as an $\A(r)$-module out of $\A(r)\mmod\A(r-k)$, in the sense that it has an increasing filtration with quotients isomorphic to $\A(r)\mmod\A(r-k)$, and present partial results towards that claim.  In addition, we prove some interesting commutation relations in the Steenrod algebra relating to representations of Steenrod Algebra elements in Wood's $Z$-basis.
\end{abstract}

@@

The structure of the Steenrod Algebra $\A$ and its various subalgebras has been the object of much study.  Consider in particular the length filtration of $\A$, with filtration quotients $L(k)$ spanned by sequences of Steenrod operations of length $k$.

@@

@@

%TODO: pictures of small ones
In particular, $\A(n)\mmod\A(n-k)$ has as representatives all elements of $\A(n)$ which cannot be formed by multiplying an element of $\A(n)$ by a nontrivial element of $\A(n-k)$.  We conjecture that $L(k)$, when considered as an $\A(n)$-module, is built up out of $\A(n)\mmod\A(n-k)$, in the sense that it admits a filtration which, except for some aperiodic parts at the beginning, has each filtration quotient isomorphic to $\A(n)\mmod\A(n-k)$.

@@

  %TODO: check this more carefully.
  First, let $y\in L(1)$, $z\in \A(n-k-1)$ with $z$ not in degree $0$, so that $yz\in M_{1,n-k-1}$.  Then by looking at degrees, $by$ is already in admissible form, so it is in $L(k+1)$, so $byz\in M_{k+1,n-k-1}$ as desired.  Since elements of the form $yz$ span $M_{1,n-k-1}$, we get that $bM_{1,n-k-1}\subset M_{k+1,n-k-1}$.

  Second, let $y\in L(k)$, $z\in \A(n-k)$ with $z$ not in degree $0$, so that $yz\in M_{k,n-k}$.  If $z\not\in\A(n-k-1)$, then $zX_{n-k}$ and thus $yzt'$ must be zero in $\A(n-k)$.  Then we may assume $z\in\A(n-k-1)$; we will show that both $yzb'$ and $yzM_{1,n-k-1}$ are contained in $M_{k+1,n-k-1}$.  Now if we write $zb'$ in admissible form, multiplying by $y$ will give an element already in admissible form by looking at degrees, so we need only show that $zb'\in M_{1,n-k-1}$.  If we multiply out using the Adem relations, this is clearly true, modulo showing that $zb'$ contains no terms of length $1$.  This is easy to show; by the Milnor basis, $z$ cannot end in a Steenrod square with degree a multiple of $2^{n-k}$, but $b'$ is a Steenrod square in a degree which is $1$ modulo $2^{n-k}$, so on multiplying the two, the length $1$ term must vanish, as desired.

@@

  First, consider the case where not all $b_i$ are in the same degree; let $b_j$ be one of the ones in the highest possible degree.  Since $\A(n)\mmod\A(n-k)$ is a Poincar\'e algebra, there exists some $y_j$ in $\A(n)\mmod\A(n-k)$ such that $y_j x_j = z$.  Then for any $b_{j'}$ in a lower degree than $b_j$, the corresponding $x_{j'}$ will be in a higher degree than $x_j$.  Then $y_j x_j = 0$, so multiplying the entire relation by $y_j$ leaves a new relation amongst only the $b_i$ in the highest possible degree.

@@

%TODO: pictures of the filtrations of L(1) and L(2)?

@@

This paper would not have been possible without the support of Mark Behrens, who proposed the problem and provided guidance along the way, and funding from the MIT Class of 1995 and Paul E. Gray Funds for UROP.

>>>>

\title{Determining the Structure of Length-$k$ Steenrod Operations as $\A(r)$-Modules}
\author{Benjamin Kraft}
\date{}

@@

\maketitle

\begin{abstract}
  Let $\A$ be the Steenrod Algebra, and let $\A(k)$ be the subalgebra generated by the $Sq^{2^i}$ for $i\leq k$.  Consider the modules $L(k)$ spanned by sequences of Steenrod operations of length $k$.  Welcher proved that $L(k)$ is a free module over $\A(k-1)$.  We are interested in finding the structure of $L(k)$ as an $\A(r)$-module for any $r$.  We conjecture that $L(k)$ is built as an $\A(r)$-module out of $\A(r)\mmod\A(r-k)$, in the sense that it has an increasing filtration with quotients isomorphic to $\A(r)\mmod\A(r-k)$, and present partial results towards that claim.  In addition, we prove some interesting commutation relations in the Steenrod algebra relating to representations of Steenrod Algebra elements in Wood's $Z$-basis.
\end{abstract}

@@

The structure of the Steenrod Algebra $\A$ and its various subalgebras has been the object of much study.  Consider in particular the length filtration of $\A$, with filtration quotients $L(k)$ spanned by sequences of Steenrod operations of length $k$.  This filtration and its relation to the sphere spectrum has been studied by a number of authors~\cite{nakaoka,welcher,welcher2}.

@@

\begin{figure}
  \begin{center}
    \includegraphics[height=1.4in]{pics/A1.pdf}
    \hspace{1in}
    \includegraphics[height=1.3in]{pics/A1-A0.pdf}
  \end{center}
  \caption{At left, a basis for the subalgebra $\A(1)$.  The dark lines represent the action of $\Sq^1$ (so for instance $\Sq^1 \cdot \Sq^2\Sq^1 = \Sq^3\Sq^1$), and the light lines represent that of $\Sq^2$.  At right, representatives for a basis of $\A(1)\mmod\A(0)$.  Again the dark lines represent the action of $\Sq^1$ and the light lines represent that of $\Sq^2$.\label{fig:A1}}
\end{figure}

@@

An example is given in Figure~\ref{fig:A1}.  In particular, $\A(n)\mmod\A(n-k)$ has as representatives all elements of $\A(n)$ which cannot be formed by multiplying an element of $\A(n)$ by a nontrivial element of $\A(n-k)$.  We conjecture that $L(k)$, when considered as an $\A(n)$-module, is built up out of $\A(n)\mmod\A(n-k)$, in the sense that it admits a filtration which, except for some aperiodic parts at the beginning, has each filtration quotient isomorphic to $\A(n)\mmod\A(n-k)$.

@@

  First, let $y\in L(1)$, $z\in \A(n-k-1)$ with $z$ not in degree $0$, so that $yz\in M_{1,n-k-1}$.  Then by looking at degrees, $by$ is already in admissible form, so it is in $L(k+1)$, so $byz\in M_{k+1,n-k-1}$ as desired.  Since elements of the form $yz$ span $M_{1,n-k-1}$, we get that $bM_{1,n-k-1}\subset M_{k+1,n-k-1}$.

  Second, let $y\in L(k)$, $z\in \A(n-k)$ with $z$ not in degree $0$, so that $yz\in M_{k,n-k}$.  If $z\not\in\A(n-k-1)$, then $zX_{n-k}$ and thus $yzt'$ must be zero in $\A(n-k)$.  Then we may assume $z\in\A(n-k-1)$; we will show that both $yzb'$ and $yzM_{1,n-k-1}$ are contained in $M_{k+1,n-k-1}$.  Now if we write $zb'$ in admissible form, multiplying by $y$ will give an element already in admissible form by looking at degrees, so we need only show that $zb'\in M_{1,n-k-1}$.  If we multiply out using the Adem relations, this is clearly true, modulo showing that $zb'$ contains no terms of length $1$.  This is easy to show; since it is in $\A(n-k-1)$, $z$ cannot end in a Steenrod square with degree a multiple of $2^{n-k}$, but $b'$ is a Steenrod square in a degree which is $1$ modulo $2^{n-k}$, so on multiplying the two, the length $1$ term must vanish, as desired.

@@

  First, consider the case where not all $b_i$ are in the same degree; let $b_j$ be one of the ones in the highest possible degree.  Since $\A(n)\mmod\A(n-k)$ is a Poincar\'e algebra, there exists some $y_j$ in $\A(n)\mmod\A(n-k)$ such that $y_j x_j = z$.  (See for example Margolis~\cite{margolis}.)  Then for any $b_{j'}$ in a lower degree than $b_j$, the corresponding $x_{j'}$ will be in a higher degree than $x_j$.  Then $y_j x_j = 0$, so multiplying the entire relation by $y_j$ leaves a new relation amongst only the $b_i$ in the highest possible degree.

@@

\begin{figure}
  \begin{center}
    \includegraphics[height=2.5in]{pics/L1-1.pdf}
  \end{center}
  \caption{The structure of $\L(1)$ as an $\A(1)$-module.  (As before, the heavy lines show the action of $\Sq^1$, while the lighter lines show the action of $\Sq^2$.)  The colors show each filtration quotient isomorphic to $\A(1)\mmod\A(0)$; the single element $\Sq^1$ is excluded from this, but would be included if we continued downwards to $\Lkm[1]$. \label{fig:L1-1}}
\end{figure}

A picture of the case of $\L(1)$ as an $\A(1)$-module is shown in Figure~\ref{fig:L1-1}.

@@

This paper would not have been possible without the support of Mark Behrens, who proposed the problem and provided guidance along the way, and sarah-marie belcastro, whose advice during the writing process was invaluable.  The project was generously funded by the MIT Class of 1995 and Paul E.~Gray Funds for UROP.
